---
title: "directed effort"
output: html_document
date: "2022-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lubridate, here, RColorBrewer, gganimate, circlize, imputeTS, lme4, MuMIn, car, DHARMa, TSA, forecast, tscount, nlme, cowplot, readxl)
```
wait, it's not in federal waters. I'll do some exploration, but I'll need to ask for federal waters estimates to make this direct comparison. 
I've reached out to Lucas again to either debug the custom data query or (maybe, if he's feeling nice) run the series in house. 

Meanwhile, quick look at trends in directed trip effort. (though this also includes people catching these species incidentally)

```{r}
bsb.1<-read_excel(here::here("directed effort","DT_Output_1981_2003_BLACKSEABASS.xls"))
bsb.2<-read_excel(here::here("directed effort","DT_Output_2004_2020_BLACKSEABASS.xls"))
bsb.effort<-rbind.data.frame(bsb.1, bsb.2)

smf.1<-read_excel(here::here("directed effort","DT_Output_1981_2003_SUMMERFLOUNDER.xls"))
smf.2<-read_excel(here::here("directed effort","DT_Output_2004_2020_SUMMERFLOUNDER.xls"))
smf.effort<-rbind.data.frame(smf.1, smf.2)

scup.1<-read_excel(here::here("directed effort","DT_Output_1981_2003_SCUP.xls"))
scup.2<-read_excel(here::here("directed effort","DT_Output_2004_2020_SCUP.xls"))
scup.effort<-rbind.data.frame(scup.1, scup.2)


short.states<-tolower(c("Massachusetts","Connecticut","Rhode Island","New York","New Jersey","Delaware","Maryland","Virginia","North Carolina"))

all.species<-rbind.data.frame(bsb.effort, smf.effort, scup.effort)

effort<-all.species%>%
  mutate(state=tolower(st_name))%>%
  filter(state%in%short.states)%>%
  select(YEAR, state, COMMON_NAME, dir_tot, pse_dt)%>%
  mutate(se=dir_tot*(pse_dt/100),
         upper_95=dir_tot+1.96*se,
         lower_95=dir_tot-1.96*se)

names(effort)<-tolower(names(effort))
```

plot effort by state and species

```{r}
palette<-brewer.pal(length(short.states), "Paired")

ggplot(effort)+
  geom_line(aes(x=year, y=dir_tot, color=state))+
  facet_grid(common_name~., scales="free_y")+
  scale_color_manual(values=palette)+
  ylab("Directed effort (trips)")+
  theme_classic()
ggsave(here::here("figures","directed.effort.png"), height=12, width=8)

```




what about trends in effort by area and mode? 

```{r}
effort.area.mode<-read_csv(here::here("data","mrip_effort_area_mode.csv"), skip=24)

effort.area<-effort.area.mode%>%
  filter(tolower(State)%in%short.states)%>%
  group_by(Year, State, `Fishing Area`)%>%
  summarize(nTrips=sum(`Angler Trips`, na.rm=T))

effort.mode<-effort.area.mode%>%
  filter(tolower(State)%in%short.states)%>%
  group_by(Year, State, `Fishing Mode`)%>%
  summarize(nTrips=sum(`Angler Trips`, na.rm=T))
```

```{r}
ggplot(effort.area)+
  geom_line(aes(x=Year, y=nTrips, color=State))+
  facet_grid(`Fishing Area`~., scales="free_y")+
  scale_color_manual(values=palette)+
  theme_classic()
ggsave(here::here("figures","effort.area.png"), height=12, width=8)
```
```{r}
ggplot(effort.mode)+
  geom_line(aes(x=Year, y=nTrips, color=State))+
  facet_grid(`Fishing Mode`~., scales="free_y")+
  scale_color_manual(values=palette)+
  theme_classic()
ggsave(here::here("figures","effort.mode.png"), height=16, width=8)
```

