---
title: "MRIP bycatch"
output: html_document
date: "2023-01-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lubridate, here, RColorBrewer, gganimate, circlize, wesanderson, ggpubr)
```

MRIP raw trip data from here: https://www.st.nmfs.noaa.gov/st1/recreational/MRIP_Survey_Data/CSV/

Skip this from now on, it's time consuming and I can just read it in. 
```{r}

 tbl<-list.files(path=here::here("data", "mrip raw"), pattern="catch*")
 
 setwd(here::here("data","mrip raw"))
 data<-sapply(tbl, read_csv, simplify=F, 
              col_types = cols(.default="c"))
 
 
 setwd("C:/Users/ashle/Dropbox/mrip_out_of_state")
 
 for(i in 1:length(data)){
    names(data[[i]])<-tolower(names(data[[i]]))
  
 }
 
 cols.data<-rep(NA, length(data))
 
 # different numbers of columns
 for(i in 1:length(data)){
   cols.data[i]<-ncol(data[[i]])
 }
 

  rbind.all.columns <- function(x, y) {
   
      x.diff <- setdiff(colnames(x), colnames(y))
      y.diff <- setdiff(colnames(y), colnames(x))
   
      x[, c(as.character(y.diff))] <- NA
   
      y[, c(as.character(x.diff))] <- NA
   
      return(rbind(x, y))
  }
  
  all.catch.data <- Reduce( rbind.all.columns, data)

```
 
 
 now trip data
 
```{r}
 tbl<-list.files(path=here::here("data", "mrip raw"), pattern="trip*")
 
 setwd(here::here("data","mrip raw"))
 data<-sapply(tbl, read_csv, simplify=F, 
              col_types = cols(.default="c"))
 
 
 setwd("C:/Users/ashle/Dropbox/mrip_out_of_state")
 
 for(i in 1:length(data)){
    names(data[[i]])<-tolower(names(data[[i]]))
  
 }
 


  rbind.all.columns <- function(x, y) {
   
      x.diff <- setdiff(colnames(x), colnames(y))
      y.diff <- setdiff(colnames(y), colnames(x))
   
      x[, c(as.character(y.diff))] <- NA
   
      y[, c(as.character(x.diff))] <- NA
   
      return(rbind(x, y))
  }
  
  all.trip.data <- Reduce( rbind.all.columns, data)

write.csv(all.catch.data, here::here("summary output","all.catch.data.csv"))
write.csv(all.trip.data, here::here("summary output","all.trip.data.csv"))
```
 
 For when this data is ready to go (note that everything is saved as a character--will have to convert as needed)
 
```{r} 
all.catch.data<-read_csv(here::here("summary output","all.catch.data.csv"))
all.trip.data<-read_csv(here::here("summary output","all.trip.data.csv"))
```
Join catch to trip data, filter to ME:NC
Filter to all trips that did not target summer flounder, black sea bass, or scup. 

Coming back to this later, make sure that all digits of PSU ID and ID code are being read in. 

```{r}
catch.trip<-left_join(all.trip.data[,c("id_code","prim1_common","prim2_common","psu_id")], all.catch.data, by=c("id_code", "psu_id"))

# psu_id and id_code together identify individual interviews


codes<-read_csv("mrip.codes.csv")%>%
  filter(`common name`%in%c("fluke","scup","black sea bass"))
fips<-read_csv(here::here("data","state.fips.csv"))%>%
  filter(stname%in%c("Maine","New Hampshire","Massachusetts","Connecticut","Rhode Island","New York","New Jersey","Delaware","Maryland","Virginia","North Carolina"))

'%!in%'<-Negate('%in%')

focal<-c("SUMMER FLOUNDER","BLACK SEA BASS","SCUP")

# estimating catch rates for focal species among trips where they were *not* specifically targeted. 

# non-targeted catch
catch.trip.f<-catch.trip%>%
  mutate(st=as.numeric(st))%>%
  filter(st%in%fips$st)%>%
  filter(prim1_common%!in%focal & prim2_common%!in%focal)%>%
  left_join(fips, by="st")
```

Estimate catch rates of black sea bass, scup, and summer flounder in each state and year on nontargeted trips. (bycatch cpue)

Keeping all modes together for now, calculating weighted mean CPUE by year and species from catch data. First step is weighted catch (data.sum below), then dividing by sum of weights across strata

```{r}
data.sum.annual<-catch.trip.f%>%
  mutate(wp_catch=as.numeric(wp_catch),
         landing=as.numeric(landing),
         release=as.numeric(release),
         year=as.numeric(year))%>%
  filter(common%in%focal)%>%
  mutate(nfishWeight=wp_catch*(landing+release))%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch=sum(nfishWeight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch2=sum(sumCatch),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumCatch3=sum(sumCatch2))%>%
  ungroup()

ggplot(data.sum.annual)+
  geom_line(aes(x=year, y=sumCatch3, color=common))+
  facet_grid(stusps~.)
```
```{r}
weight.sum.annual<-catch.trip.f%>%
    mutate(wp_catch=as.numeric(wp_catch),
         landing=as.numeric(landing),
         release=as.numeric(release),
         year=as.numeric(year),
         wp_int=as.numeric(wp_int))%>%
  filter(common%in%focal)%>%
  mutate(weight=wp_int)%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight=sum(weight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight2=sum(sumWeight),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumWeight3=sum(sumWeight2))%>%
  ungroup()
```


Now bind those dfs together, calculate CPUE
```{r}
data.bind<-left_join(data.sum.annual, weight.sum.annual[,c("year","stusps","common","sumWeight3")], by=c("year","stusps","common"))

data.bind$cpue<-data.bind$sumCatch3/data.bind$sumWeight3

data.bind$stusps<-factor(data.bind$stusps, levels=c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC"))
```


```{r}
palette<-wes_palette("BottleRocket2",3,"discrete")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9")

ggplot(data.bind)+
  geom_line(aes(x=year, y=cpue, color=common))+
  geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
  geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
  geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  ggtitle("Bycatch cpue")+
  scale_color_manual(values=cbbPalette)+
  facet_grid(stusps~.)+
  theme_bw()
ggsave(here::here("figures","bycatch.cpue.png"), height=12, width=4)
```
Hypothesis: as RHL goes down (and regulations become more restrictive), bycatch of focal species goes up. 

This is a catch rate, so to estimate total annual bycatch, I need nontargeted fishing effort. This can be total fishing effort (all species) - targeted fishing effort for these species. 


```{r}
rhl<-read_csv(here::here("state regs","rhl.csv"))%>%
  mutate(species=tolower(species))

bycatch.cpue<-data.bind%>%
  mutate(species=tolower(common))%>%
  left_join(rhl, by=c("species","year"))
```


I should remember to compare bycatch to targeted cpue

So next few chunks are estimating total CPUE

```{r}


catch.trip.total<-catch.trip%>%
  filter(st%in%fips$st)%>%
  mutate(st=as.numeric(st))%>%
  # comparing bycatch cpue instead to total cpue to see if psu problems were caused by target species filtering
  #filter(prim1_common%in%focal | prim2_common%in%focal)%>%
  left_join(fips, by="st")

data.sum.annual.total<-catch.trip.total%>%
  filter(common%in%focal)%>%
  mutate(wp_catch=as.numeric(wp_catch),
         landing=as.numeric(landing),
         release=as.numeric(release))%>%
  mutate(nfishWeight=wp_catch*(landing+release))%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch=sum(nfishWeight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch2=sum(sumCatch),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumCatch3=sum(sumCatch2))%>%
  ungroup()
  
# ah, strat_id is a subset of psu, not the other way around. 
```



```{r}
weight.sum.annual.total<-catch.trip.total%>%
  filter(common%in%focal)%>%
  mutate(wp_int=as.numeric(wp_int))%>%
  mutate(weight=wp_int)%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight=sum(weight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight2=sum(sumWeight),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumWeight3=sum(sumWeight2))%>%
  ungroup()
```


Now bind those dfs together, calculate CPUE
```{r}
data.bind.total<-left_join(data.sum.annual.total, weight.sum.annual.total[,c("year","stusps","common","sumWeight3")], by=c("year","stusps","common"))


data.bind.total$cpue<-data.bind.total$sumCatch3/data.bind.total$sumWeight3
data.bind.total$year<-as.numeric(data.bind.total$year)

data.bind.total$stusps<-factor(data.bind.total$stusps, levels=c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC"))


```


```{r}

ggplot(data.bind.total)+
  geom_line(aes(x=year, y=cpue, color=common))+
   geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
   geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
   geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  ggtitle("Total cpue")+
  scale_color_manual(values=cbbPalette)+
  facet_grid(stusps~.)+
  theme_bw()
ggsave(here::here("figures","total.cpue.png"), height=12, width=4)
```
Add total cpue to to bycatch cpue and rhl df

```{r}
bycatch.cpue$cpueBycatch<-bycatch.cpue$cpue
bycatch.cpue$rhl<-bycatch.cpue$`quota (millions of pounds)`
data.bind.total$cpueTotal<-data.bind.total$cpue

all.cpue.est<-left_join(data.bind.total[,c("year","common","stusps","cpueTotal")], bycatch.cpue[,c("year","common","stusps","cpueBycatch")], 
                        by=c("year","common","stusps"))

# ugh, just join on RHL separately. 

rhl$common<-toupper(rhl$species)
rhl$rhl<-rhl$`quota (millions of pounds)`

all.cpue.est.2<-left_join(all.cpue.est, rhl[,c("common","year","rhl")], by=c("common","year"))

all.cpue.est<-all.cpue.est.2

```

Problem with this join--resulting dataframe has NA values for some species and years' RHL. This is because the states and years were missing from bycatch.cpue. (because not all states and years had quantified catch)



And last target cpue


```{r}


catch.trip.target<-catch.trip%>%
  mutate(st=as.numeric(st))%>%
  filter(st%in%fips$st)%>%
  filter(prim1_common%in%focal | prim2_common%in%focal)%>%
  left_join(fips, by="st")

data.sum.annual.target<-catch.trip.target%>%
  filter(common%in%focal)%>%
  mutate(wp_catch=as.numeric(wp_catch),
         landing=as.numeric(landing),
         release=as.numeric(release))%>%
  mutate(nfishWeight=wp_catch*(landing+release))%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch=sum(nfishWeight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumCatch2=sum(sumCatch),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumCatch3=sum(sumCatch2))%>%
  ungroup()
  
# ah, strat_id is a subset of psu, not the other way around. 
```



```{r}
weight.sum.annual.target<-catch.trip.target%>%
  filter(common%in%focal)%>%
  mutate(weight=as.numeric(wp_int))%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight=sum(weight),
            year=unique(year),
            wave=unique(wave),
            stusps=unique(stusps),
            psu_id=unique(psu_id))%>%
  ungroup()%>%
  group_by(strat_id, common)%>%
  summarize(sumWeight2=sum(sumWeight),
            year=unique(year),
            stusps=unique(stusps))%>%
  ungroup()%>%
  group_by(year, common, stusps)%>%
  summarize(sumWeight3=sum(sumWeight2))%>%
  ungroup()
```


Now bind those dfs together, calculate CPUE
```{r}
data.bind.target<-left_join(data.sum.annual.target, weight.sum.annual.target[,c("year","stusps","common","sumWeight3")], by=c("year","stusps","common"))

data.bind.target$cpue<-data.bind.target$sumCatch3/data.bind.target$sumWeight3

data.bind.target$stusps<-factor(data.bind.target$stusps, levels=c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC"))

data.bind.target$year<-as.numeric(data.bind.target$year)


```


```{r}

ggplot(data.bind.target)+
  geom_line(aes(x=year, y=cpue, color=common))+
  geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
  geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
  geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  ggtitle("Target cpue")+
  scale_color_manual(values=cbbPalette)+
  facet_grid(stusps~.)+
  theme_bw()
ggsave(here::here("figures","target.cpue.png"), height=12, width=4)
```

Visually, it looks to me like targeted cpue goes down after implementation of RHLs (not very surprising), but bycatch cpue stays about the same. 

```{r}
data.bind.target$cpueTarget<-data.bind.target$cpue

all.cpue<-left_join(all.cpue.est, data.bind.target[,c("year","common","stusps","cpueTarget")], by=c("year","common","stusps"))
```


Ok, the problem is that I need directed fishing effort by year and state for all species that are not summer flounder, black sea bass, or scup. I have a request in for directed effort towards these three species. I can subtract that effort from total effort to estimate total bycatch and its relationship to rhl. 

as of 6/15 I now have that data. (previous problems with estimates--identical estimates for each species)

```{r}
directed.bsb<-read_csv(here::here("directed effort","updated_estimates","BLACKSEABASS.csv"))%>%
  mutate(species="BLACK SEA BASS")
directed.scup<-read_csv(here::here("directed effort","updated_estimates","SCUP.csv"))%>%
  mutate(species="SCUP")
directed.smf<-read_csv(here::here("directed effort","updated_estimates","SUMMERFLOUNDER.csv"))%>%
  mutate(species="SUMMER FLOUNDER")


# codes for state, mode, area_x
fips<-read_csv(here::here("data","state.fips.csv"))%>%
  mutate(st=as.character(st))



directed<-rbind.data.frame(directed.bsb, directed.scup, directed.smf)%>%
  # problem: some one-digit FIP codes, so need to check for length of my_dom_id
  mutate(my_dom_id=as.character(my_dom_id),
         length=str_length(my_dom_id),
         year=as.numeric(str_sub(my_dom_id,1,4)),
         st=ifelse(length==7, str_sub(my_dom_id, 5,5),
                      str_sub(my_dom_id, 5,6)),
         mode=ifelse(length==7, str_sub(my_dom_id, 6,6),
                     str_sub(my_dom_id, 7,7)),
         area_x=ifelse(length==7, str_sub(my_dom_id, 7,7),
                       str_sub(my_dom_id, 8,8)))%>%
  mutate(mode=recode(mode,
                         "3"="shore",
                         "4"="forHire",
                         "6"="forHire",
                         "7"="privateBoat",
                         "5"="forHire"
                         ))%>%
  mutate(area=recode(area_x, 
                         "1"="stateWaters",
                         "2"="federalWaters",
                         "5"="inland",
                         "3"="WFL_less_10",
                         "4"="WFL_more_10"))%>%
  left_join(fips, by="st")%>%
  select(!c("Domain","Variable Name", "st", "area_x"))


# totally aggregate in federal waters, separate only by year and state
directed.agg<-directed%>%
  filter(area=="federalWaters")%>%
  group_by(year, species, stusps)%>%
  summarize(directedEffort=sum(Sum),
            directedEffort.se=sqrt(sum(VarSum)),
            CV=directedEffort.se/directedEffort,
            stname=unique(stname))%>%
  ungroup()%>%
  filter(stname%in%c("Maine","New Hampshire","Massachusetts","Connecticut","Rhode Island","New York","New Jersey","Delaware","Maryland","Virginia","North Carolina"))%>%
  mutate(stusps=factor(stusps, levels=c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC")),
         year=as.numeric(year))

directed.agg.df<-as.data.frame(directed.agg)



ggplot(directed.agg, aes(x=year, y=directedEffort, color=species, group=species))+
  geom_line()+
   geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
   geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
   geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  ggtitle("Directed effort")+
  scale_color_manual(values=cbbPalette)+
  facet_grid(stusps~.)+
  theme_bw()
ggsave(here::here("figures","directed.effort.png"), height=12, width=4)


bsb<-filter(directed.agg, species=="BLACK SEA BASS")
smf<-filter(directed.agg, species=="SUMMER FLOUNDER")
scup<-filter(directed.agg, species=="SCUP")

ggplot(scup, aes(x=year, y=directedEffort))+
  geom_line(aes(color=stname))
ggplot(bsb, aes(x=year, y=directedEffort))+
  geom_line(aes(color=stname))
ggplot(smf, aes(x=year, y=directedEffort))+
  geom_line(aes(color=stname))

```

How did total catch of bsb, scup, summer flounder change in non-targeted trips?

```{r}
directed.all<-directed%>%
  filter(area=="federalWaters")%>%
  filter(stusps%in%c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC"))%>%
  group_by(year, stusps, species)%>%
  summarize(directedEffort=sum(Sum))%>%
  ungroup()%>%
  mutate(state=stusps)

ggplot(directed.all)+
  geom_line(aes(x=year, y=directedEffort, color=species))+
  facet_grid(stusps~.)+
  scale_color_manual(values=cbbPalette)+
  theme_bw()

directed.bsb<-filter(directed.all, species=="BLACK SEA BASS")%>%
  rename("nTrips.bsb"=directedEffort)
directed.smf<-filter(directed.all, species=="SUMMER FLOUNDER")%>%
  rename("nTrips.smf"=directedEffort)
directed.scup<-filter(directed.all, species=="SCUP")%>%
  rename("nTrips.scup"=directedEffort)

total.effort<-read_csv(here::here("data","mrip estimates","fed_eez_mrip_effort_series_modes_combined.csv"), skip=24)%>%
  # replace '.' with NA
  mutate(nTrips=as.numeric(na_if(`Angler Trips`,".")),
         CV=as.numeric(na_if(`PSE`,"."))/100,
         nTrips.se=CV*as.numeric(nTrips),
         year=as.numeric(Year),
         state=State,
         state=recode(state, 
                      "MAINE"="ME",
                      "NEW HAMPSHIRE"="NH",
                      "MASSACHUSETTS"="MA",
                      "RHODE ISLAND"="RI",
                      "CONNECTICUT"="CT",
                      "NEW YORK"="NY",
                      "NEW JERSEY"="NJ",
                      "DELAWARE"="DE",
                      "MARYLAND"="MD",
                      "VIRGINIA"="VA",
                      "NORTH CAROLINA"="NC"))%>%
  filter(state%in%c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC"))


# join total effort to directed effort (total effort estimates will repeat--joining by year and state (not species))

directed.total<-left_join(directed.all, total.effort[,c("year","state","nTrips")], by=c("year","state"))%>%
  mutate(nonDirectedEffort=nTrips-directedEffort)

# now join effort tp cpue df

all.cpue$species<-all.cpue$common
all.cpue$state<-all.cpue$stusps

all.cpue.effort<-left_join(all.cpue, directed.total, by=c("year","state","species"))%>%
  mutate(catchTotal=cpueTotal*nTrips,
         catchDirected=cpueTarget*directedEffort,
         catchBycatch=cpueBycatch*nonDirectedEffort,
         state=factor(state, levels=c("ME","NH","MA","RI","CT","NY","NJ","DE","MD","VA","NC")))

```

Time for graphs

```{r}
# summer flounder first

ggplot(filter(all.cpue.effort, species=="SUMMER FLOUNDER"))+
  geom_line(aes(x=year, y=catchTotal), linetype="dotted")+
  geom_line(aes(x=year, y=catchDirected), linetype="dashed")+
  geom_line(aes(x=year, y=catchBycatch))+
  facet_grid(state~.)+
  theme_bw()

```

```{r}
ggplot(filter(all.cpue.effort, species=="BLACK SEA BASS"))+
  geom_line(aes(x=year, y=catchTotal), linetype="dotted")+
  geom_line(aes(x=year, y=catchDirected), linetype="dashed")+
  geom_line(aes(x=year, y=catchBycatch))+
  facet_grid(state~.)+
  theme_bw()

```
Bycatch is higher than directed catch in NJ at least. hard to tell in other states.

```{r}
ggplot(filter(all.cpue.effort, species=="SCUP"))+
  geom_line(aes(x=year, y=catchTotal), linetype="dotted")+
  geom_line(aes(x=year, y=catchDirected), linetype="dashed")+
  geom_line(aes(x=year, y=catchBycatch))+
  facet_grid(state~.)+
  theme_bw()


```

Response of fishing effort to RHL? 

```{r}
ggplot(all.cpue.effort)+
  geom_point(aes(x=rhl, y=nTrips, color=species))+
  scale_color_manual(values=cbbPalette)+
  theme_bw()

# aggregate across states

effort.agg<-all.cpue.effort%>%
  filter(year<2019)%>%
  group_by(year, species)%>%
  summarize(rhl=unique(rhl),
            nTrips=sum(nTrips, na.rm=T),
            directedTrips=sum(directedEffort, na.rm=T),
            nonDirectedTrips=sum(nonDirectedEffort, na.rm=T))%>%
  ungroup()

total.effort<-ggplot(effort.agg)+
  geom_point(aes(x=rhl, y=nTrips, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlab("Recreational harvest limit")+
  ylab("Annual total fishing trips")+
  theme_bw()+
  theme(legend.position="none")

```
Fairly clear relationship between total effort and black sea bass RHL. (which is not reflected in directed effort! weird.)

Directed effort in theory shoudl be clearer

```{r}
directed.effort<-ggplot(effort.agg)+
  geom_point(aes(x=rhl, y=directedTrips, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlab("Recreational harvest limit")+
  ylab("Annual directed fishing trips")+
  theme_bw()+
  theme(legend.position="none")


```
Weird, it's not very. Scup has the clearest relationship, bsb and smf are just blobs. 

species specific linear regressions

```{r}
smf.effort<-lm(log(directedTrips)~rhl, filter(effort.agg, species=="SUMMER FLOUNDER"))
summary(smf.effort)
```

```{r}
bsb.effort<-lm(log(directedTrips)~rhl, filter(effort.agg, species=="BLACK SEA BASS"))
summary(bsb.effort)

```
```{r}
scup.effort<-lm(log(directedTrips)~rhl, filter(effort.agg, species=="SCUP"))
summary(scup.effort)

```
No significant relationships between directed effort and rhl for specific species

Non-directed effort--in theory it would go up as anglers switch to other species at low RHL

Test this: 

```{r}
effort.mod<-lm(directedTrips~nTrips+rhl+nTrips*rhl, filter(effort.agg, species=="SCUP"))
plot(effort.mod)
summary(effort.mod)
```
No, doesn't really hold up. 

```{r}
effort.mod.red<-lm(directedTrips~nTrips+rhl, filter(effort.agg, species=="SCUP"))
plot(effort.mod.red)
summary(effort.mod.red)
```


```{r}
nondirected.effort<-ggplot(effort.agg)+
  geom_point(aes(x=rhl, y=nonDirectedTrips, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlab("Recreational harvest limit")+
  ylab("Annual non-directed effort")+
  theme_bw()
legend<-get_legend(nondirected.effort)

nondirected.effort<-ggplot(effort.agg)+
  geom_point(aes(x=rhl, y=nonDirectedTrips, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlab("Recreational harvest limit")+
  ylab("Annual non-directed effort")+
  theme_bw()+
  theme(legend.position="none")

```

```{r}
library(cowplot)
plot_grid(total.effort, directed.effort, nondirected.effort, legend, rel_widths=c(3,3,3,1),nrow=1)
ggsave(here::here("figures","effort.vs.rhl.png"), height=4, width=12)

```


Now aggregate bycatch and investigate relationship with RHL

Replace NA RHL with 99? 

```{r}
bycatch.agg<-all.cpue.effort%>%
  filter(year<2020)%>%
  group_by(year, species)%>%
  summarize(catchBycatch=sum(catchBycatch, na.rm=T),
            rhl=unique(rhl))%>%
  ungroup()%>%
  mutate(rhl=ifelse(is.na(rhl), 99, rhl))

ggplot(bycatch.agg)+
  geom_point(aes(x=rhl, y=catchBycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  theme_bw()
```
Try taking out years without RHL

```{r}
bycatch<-ggplot(bycatch.agg)+
  geom_point(aes(x=rhl, y=catchBycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlim(0,15)+
  xlab("Recreational harvest limit")+
  ylab("Annual bycatch estimate")+
  theme_bw()+
  theme(legend.position="none")

lm.species<-lm(log(catchBycatch)~rhl+species, filter(bycatch.agg, !is.na(rhl)))
plot(lm.species)
summary(lm.species)

```
If there's a pattern there, it's really noisy. Bycatch as proportion of total catch?

```{r}
bycatch.prop<-all.cpue.effort%>%
  filter(year<2020)%>%
  group_by(year, species)%>%
  summarize(catchBycatch.agg=sum(catchBycatch, na.rm=T),
            catchTotal.agg=sum(catchTotal, na.rm=T),
            rhl=unique(rhl))%>%
  ungroup()%>%
  mutate(propBycatch=catchBycatch.agg/catchTotal.agg)

ggplot(bycatch.prop)+
  geom_point(aes(x=rhl, y=propBycatch, color=species))+
 # xlim(0,15)+
  scale_color_manual(values=cbbPalette)+
  theme_bw()
ggsave(here::here("figures","prop.bycatch.vs.rhl.png"), height=12, width=6)

```
I do see a small positive effect of RHL on summer flounder log bycatch. 

```{r}
lm.rhl.smf<-lm(log(catchBycatch)~rhl, filter(all.cpue.effort, species=='SUMMER FLOUNDER' & !is.na(rhl)))
plot(lm.rhl.smf)
summary(lm.rhl.smf)
```
```{r}
lm.rhl.bsb<-lm(log(catchBycatch)~rhl, filter(all.cpue.effort, species=='BLACK SEA BASS' & !is.na(rhl)))
plot(lm.rhl.bsb)
summary(lm.rhl.bsb)

```
```{r}
lm.rhl.scup<-lm(log(catchBycatch)~rhl, filter(all.cpue.effort, species=='SCUP' & !is.na(rhl)))
plot(lm.rhl.scup)
summary(lm.rhl.scup)

```
OK, when we look at all species together, species/year combinations with lower RHLs have higher bycatch. But separated by species, don't see the same pattern. For summer flounder, a small positive effect of RHL on bycatch (so lower RHL associated with lower bycatch)

What about by proportion of total catch? 

```{r}
lm.prop.smf<-lm(propBycatch~rhl, filter(bycatch.prop, species=="SUMMER FLOUNDER" & !is.na(rhl)))
plot(lm.prop.smf)
summary(lm.prop.smf)
```
```{r}
lm.prop.bsb<-lm(propBycatch~rhl, filter(bycatch.prop, species=="BLACK SEA BASS" & !is.na(rhl)))
plot(lm.prop.bsb)
summary(lm.prop.bsb)

```
```{r}
lm.prop.scup<-lm(propBycatch~rhl, filter(bycatch.prop, species=="SCUP" & !is.na(rhl)))
plot(lm.prop.scup)
summary(lm.prop.scup)

```
What if I don't aggregate by state? (state random effect?)

```{r}
ggplot(filter(all.cpue.effort, !is.na(rhl)))+
  geom_point(aes(x=rhl, y=catchBycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  theme_bw()

library(lme4)

m.mod<-lmer(log(catchBycatch)~rhl +  (species|state), filter(all.cpue.effort, !is.na(rhl)))
plot(resid(m.mod))
qqnorm(resid(m.mod))
summary(m.mod)

```


Plotting total and directed catch

```{r}
catch.agg<-all.cpue.effort%>%
  filter(year<2020)%>%
  group_by(year, species)%>%
  summarize(catchBycatch=sum(catchBycatch, na.rm=T),
            catchTotal=sum(catchTotal, na.rm=T),
            catchDirected=sum(catchDirected, na.rm=T),
            rhl=unique(rhl))%>%
  ungroup()%>%
  mutate(rhl=ifelse(is.na(rhl), 99, rhl))

bycatch<-ggplot(catch.agg)+
  geom_point(aes(x=rhl, y=catchBycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlim(0,15)+
  xlab("Recreational harvest limit")+
  ylab("Annual bycatch estimate")+
  theme_bw()+
  theme(legend.position="none")

directed<-ggplot(catch.agg)+
  geom_point(aes(x=rhl, y=catchDirected, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlim(0,15)+
  xlab("Recreational harvest limit")+
  ylab("Annual directed catch estimate")+
  theme_bw()+
  theme(legend.position="none")

total<-ggplot(catch.agg)+
  geom_point(aes(x=rhl, y=catchTotal, color=species))+
  scale_color_manual(values=cbbPalette)+
  xlim(0,15)+
  xlab("Recreational harvest limit")+
  ylab("Annual total catch estimate")+
  theme_bw()+
  theme(legend.position="none")

plot_grid(total, directed, bycatch, legend, nrow=1, rel_widths=c(3,3,3,1))
ggsave(here::here("figures","catch.vs.rhl.png"), height=4, width=12)
```
time series
```{r}
ggplot(filter(all.cpue.effort, year<2020))+
  geom_line(aes(x=year, y=catchBycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  facet_grid(state~.)+
  ylab("Annual bycatch estimate")+
  geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
  geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
  geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  theme_bw()
ggsave(here::here("figures","bycatch.estimate.png"), height=12, width=6)
  

```

Divide instead by pre/post RHL? 

```{r}
all.cpue.rhl.agg<-all.cpue.effort%>%
  filter(year<2020)%>%
  mutate(rhl.ind=ifelse(is.na(rhl), 0, 1),
         rhl.ind=as.factor(rhl.ind))%>%
  ungroup()
ggplot(all.cpue.rhl.agg)+
  geom_boxplot(aes(x=rhl.ind, y=directedEffort))+
  facet_grid(species~.)+
  theme_bw()
ggsave(here::here("figures","boxplot.directed.effort.rhl.png"), height=12, width=6)

ggplot(all.cpue.rhl.agg)+
  geom_boxplot(aes(x=rhl.ind, y=catchBycatch))+
  facet_grid(species~.)+
  theme_bw()
ggsave(here::here("figures","boxplot.bycatch.rhl.png"), height=12, width=6)
```
eh, no obvious change

```{r}
data.to.send<-all.cpue.effort%>%
  select(!c(stusps.x, stusps.y))

write.csv(data.to.send, here::here("summary output","catch.and.effort.estimates.MRIP.csv"))
```

aggregating all states

```{r}
agg.catch<-all.cpue.effort%>%
  filter(year<2020)%>%
  group_by(year, species)%>%
  summarize(bycatch=sum(catchBycatch, na.rm=T),
            rhl=unique(rhl))

ggplot(agg.catch)+
  geom_line(aes(x=year, y=bycatch, color=species))+
  scale_color_manual(values=cbbPalette)+
  geom_vline(xintercept=1993, color="#56B4E9", linetype="dashed")+
  geom_vline(xintercept=1998, color="#000000", linetype="dashed")+
  geom_vline(xintercept=1997, color="#E69F00", linetype="dashed")+
  ylab("Annual bycatch estimate")+
  theme_bw()
ggsave(here::here("figures","aggregate.bycatch.png"), height=4, width=6)
```


Binning data before and after RHL

```{r}
ggplot(all.cpue.rhl.agg)+
  geom_boxplot(aes(x=rhl.ind, y=cpueBycatch))+
  facet_grid(species~.)+
  theme_bw()

```

So there was no clear linear relationship between RHL and fishing effort or bycatch. I'm going to instead try a time series approach to see if introducing an RHL was an intervention that produced a permanent constant change, temporary constant change, gradual change to new mean, or initial change that gradually reverted. 

Starting with aggregate effect across all states, and potentially slicing it from there. 

let's take all.cpue.effort and aggregate it

```{r}
agg.ts.data<-all.cpue.effort%>%
  filter(year<2020)%>%
  group_by(year, species)%>%
  summarize(directedEffort=sum(directedEffort, na.rm=T),
            totalEffort=sum(nTrips, na.rm=T),
            nonDirectedEffort=sum(nonDirectedEffort, na.rm=T),
            catchTotal=sum(catchTotal, na.rm=T),
            catchDirected=sum(catchDirected, na.rm=T),
            catchBycatch=sum(catchBycatch, na.rm=T),
            rhl=unique(rhl))%>%
  ungroup()%>%
  mutate(rhl.ind=ifelse(is.na(rhl), 0, 1))
  
```


