---
title: "data wrangling"
output: html_document
date: "2022-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lubridate, here)
```

MRIP raw trip data from here: https://www.st.nmfs.noaa.gov/st1/recreational/MRIP_Survey_Data/CSV/

```{r}



tbl<-list.files(path=here::here("data", "mrip raw"), pattern="trip*")

setwd(here::here("data","mrip raw"))
data<-sapply(tbl, read_csv, simplify=F)


setwd("C:/Users/ashle/Dropbox/MRIP out of state/mrip_out_of_state")

for(i in 1:length(data)){
   names(data[[i]])<-tolower(names(data[[i]]))
 
}


#all.data<-do.call("rbind", data)

 rbind.all.columns <- function(x, y) {
  
     x.diff <- setdiff(colnames(x), colnames(y))
     y.diff <- setdiff(colnames(y), colnames(x))
  
     x[, c(as.character(y.diff))] <- NA
  
     y[, c(as.character(x.diff))] <- NA
  
     return(rbind(x, y))
 }
```
 
 
 
```{r} 
 
all.data <- Reduce( rbind.all.columns, data)

write.csv(all.data, here::here("summary output", "all.trip.data.csv"))
```

If I don't want to do all of the above from scratch in the future: 

```{r}
all.data<-read_csv(here::here("summary output","all.trip.data.csv"))
```


```{r}
names(all.data)
```

site ID codes are not shared publically, so for now I will get centroid lat longs for county of intercept and county of residence.

county FIPS (and other cool spatial data) https://github.com/btskinner/spatial
```{r}
fips.convert<-read_csv(here::here("data","county_centers.csv"))

# clon10 is spatial center of county in 2010 census

fips<-fips.convert%>%
  select(fips, clat10, clon10)%>%
  mutate(fips=as.character(fips),
         # fips string needs to be same length as concatenated fips below in data.select; tripped me up before
         fips=str_pad(fips, 5, "left", "0"))%>%
  rename(cnty_fips=fips,
         latInt=clat10,
         longInt=clon10)

fips.res<-fips%>%
  rename(cnty_res_fips=cnty_fips,
         latRes=latInt,
         longRes=longInt)


```


```{r}
# I need to make a new column that combines state and county fips
data.select<-all.data%>%
  select(area_x, boat_hrs, cnty, cnty_res, dist, gear, hrs_dtd, hrsf, intsite, kod, mode_f, month, prim1_common, prim2_common, psu_id, id_code, st, st_res, strat_id, time, wave, year, zip)%>%
  mutate(cnty_res_fips=paste0(str_pad(as.character(st_res), 2, "left", "0"), str_pad(as.character(cnty_res), 3, "left", "0")),
         cnty_fips=paste0(str_pad(as.character(st), 2, "left", "0"), str_pad(as.character(cnty), 3, "left", "0")))%>%
  left_join(fips.res, by="cnty_res_fips")%>%
  left_join(fips, by="cnty_fips")%>%
  # remove the 48000 entries with missing coordinates
  filter(!is.na(latInt) & !is.na(latRes))%>%
  mutate(outOfState=ifelse(st==st_res, 0, 1),
         fromNorth=ifelse(latRes>latInt, 1, 0),
         fromSouth=ifelse(latRes<latInt, 1, 0))


data.outOfState<-filter(data.select, outOfState==1)  

hist(data.outOfState$fromNorth)
```
```{r}
hist(data.outOfState$fromSouth)
```
Huh. And what if I restrict it to mid-atlantic?

States: North Carolina, Virginia, Maryland, New Jersey, Delaware, Pennsylvania, New York, Connecticut, Rhode Island, Massachusetts, New Hampshire, and Maine
```{r}
midatl.outOfState<-filter(data.outOfState,
                          st%in%c(37, 51, 24, 34, 10, 42, 36, 9, 44, 25, 33, 23))

# that's cumbersome, just join the state/fips csv

state.fips<-read_csv(here::here("data","state.fips.csv"))

midatl.visitors<-left_join(midatl.outOfState, state.fips[,c("st", 'stusps')], by="st")

state.fips$st_res<-state.fips$st
state.fips$stusps_res<-state.fips$stusps

midatl.visitors<-left_join(midatl.visitors, state.fips[,c("st_res","stusps_res")], by="st_res")

ggplot(midatl.visitors)+
  geom_bar(aes(x=stusps, fill=stusps_res ), position="stack")
```
oof that's ugly. 

```{r}
midatl.visitors$fromWhere<-ifelse(midatl.visitors$fromNorth==1, "north", 
                                  ifelse(midatl.visitors$fromSouth==1,"south","NA"))

midatl.visitors$stusps<-factor(midatl.visitors$stusps, levels=c("NC","VA","MD","DE","NJ","NY","CT","RI","MA","NH","ME"))

ggplot(midatl.visitors)+
  geom_bar(aes(x=stusps, fill=fromWhere), position="stack")
```
Right, so clearly the further north you are, the more likely out of staters are to come from the south. That's obvious. Over time for each state, how has that number changed? 


To do: 
- match residence ZIP code to centroid lat long and US state
- compare residence lat long to access point lat long, classify as traveling north or south
- by state and year, summarize number of intercepts of out of state anglers traveling from 1) north and 2) south
- by state and year, summarize number of intercepts of all anglers traveling from 1) north and 2) south
- by state and year for summer flounder, black sea bass, and scup, summarize number of intercepts of out of state anglers traveling from north and south. 
- fit time series models: categorical year effect, linear effect, seasonality (categorical by wave, probably)
- maybe if I hate myself: collect state regulations to add as covariates. 
